---
title: "Advent of code 2020 day 8"
description: |
  --- Day 8: Handheld Halting ---
author:
  - name: Philippe Massicotte
date: 12-14-2020
output:
  distill::distill_article:
    self_contained: false
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(here)
```

https://adventofcode.com/2020/day/8

## Read the input data

First lets read the data and turn it into a tibble.

```{r}
file <- here("data", "input", "day8.txt")

df <- file %>%
  read_lines() %>%
  enframe(name = "operation_id", value = "operation")

df %>%
  rmarkdown::paged_table()
```

Then separate the `operation` column into actual operation and the associated numerical value.

```{r}
df <- df %>%
  separate(
    operation,
    into = c("operation", "value"),
    sep = 3,
    convert = TRUE
  )

df %>%
  rmarkdown::paged_table()
```

## Part 1

Add some columns to the tibble.

```{r}
df <- df %>%
  mutate(visited = FALSE, .after = "operation_id") %>%
  mutate(accumulator = 0) %>%
  mutate(cur_operation_id = 1)
```


```{r}
get_new_index <- function(df) {
  cur_operation_id <- unique(df$cur_operation_id)
  operation <- df$operation[cur_operation_id]

  if (operation == "nop") {
    df$visited[cur_operation_id] <- TRUE
    df$cur_operation_id <- df$cur_operation_id + 1
  } else if (operation == "acc") {
    df$accumulator <- df$accumulator + df$value[cur_operation_id]
    df$visited[cur_operation_id] <- TRUE
    df$cur_operation_id <- df$cur_operation_id + 1
  } else if (operation == "jmp") {
    df$visited[cur_operation_id] <- TRUE
    df$cur_operation_id <- df$cur_operation_id + df$value[cur_operation_id]
  }

  # If the next operation id is greater than the number of rows in the df
  if (unique(df$cur_operation_id) > nrow(df)) {
    df$cur_operation_id <- nrow(df)
    return(df)
  }

  return(df)
}
```

```{r}
while (!df$visited[unique(df$cur_operation_id)]) {
  df <- get_new_index(df)
}

unique(df$accumulator)
```

## Part 2

*Reset* the tibble for the next part.

```{r}
df <- df %>%
  mutate(visited = FALSE, .after = "operation_id") %>%
  mutate(accumulator = 0) %>%
  mutate(cur_operation_id = 1)
```

Here I am going to use a brute force approach to find out when the last operation is executed. The number of simulations to perform corresponds to the number of `nop` and `jmp` found in the data.

```{r}
res <- df %>%
  mutate(n = operation_id) %>%
  mutate(n = ifelse(operation %in% c("nop", "jmp"), n, NA))

# How many simulations have to be performed
n <- res %>%
  pull(n) %>%
  na.omit() %>%
  length()

n

# Which rows of the df that contain nop or jmp have to be swaped
rows_to_modify <- res %>%
  pull(n) %>%
  na.omit()

head(rows_to_modify)

res <- res %>%
  select(-n) %>%
  crossing(simulation_id = 1:n) %>%
  group_nest(simulation_id, keep = TRUE) %>%
  mutate(rows_to_modify) %>%
  mutate(data = map2(data, rows_to_modify, function(data, rows_to_modify) {
    data$operation[rows_to_modify] <-
      ifelse(data$operation[rows_to_modify] == "nop", "jmp", "nop")

    return(data)
  }))

res %>%
  rmarkdown::paged_table()
```

We can see that there are `r nrow(res)` simulations to be done.

```{r simulation_day08, cache=TRUE}
loop <- function(df) {
  while (!df$visited[unique(df$cur_operation_id)]) {
    df <- get_new_index(df)
  }

  return(df)
}

res <- res %>%
  mutate(res = map(data, loop))
```

```{r}
res %>%
  select(-simulation_id) %>%
  unnest(res) %>%
  filter(cur_operation_id >= 629) %>%
  pull(accumulator) %>%
  unique()
```
