---
title: "Advent of code 2020 day 9"
description: |
  --- Day 9: Encoding Error ---
author:
  - name: Philippe Massicotte
date: 12-15-2020
output:
  distill::distill_article:
    self_contained: false
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(here)
library(slider)
```

https://adventofcode.com/2020/day/9

## Read the input data

First, let's read the data and turn it into a tibble.

```{r}
file <- here("data", "input", "day9.txt")

df <- file %>%
  read_lines() %>%
  enframe(name = "number_id", value = "number") %>%
  mutate(number = parse_number(number))

df %>%
  head(100) %>% 
  rmarkdown::paged_table()
```

## Part 1

Create vectors with the previous 25 observations.

```{r}

preamble_length <- 25

df <- df %>%
  mutate(data = slide(number,
    ~ .x[1:preamble_length],
    .before = preamble_length,
    .complete = TRUE
  ))

df %>%
  head(100) %>% 
  rmarkdown::paged_table()
```

Remove cases where there were not enough observations to extract 25 previous values.

```{r}
df <- df %>%
  filter(!map_lgl(data, is.null))

df %>% 
  head(100) %>% 
  rmarkdown::paged_table()
```

Find all the combinations of 2 numbers and calculate the sum of these numbers.

```{r}
df <- df %>%
  mutate(total = map(data, function(data) {
    total <- combn(data, m = 2) %>%
      t() %>%
      rowSums()

    return(total)
  })) %>%
  unnest(total)

df %>% 
  head(100) %>% 
  rmarkdown::paged_table()
```

Finally, find the number where two numbers sum up to this number.

```{r}
invalid_number <- df %>%
  group_by(number_id) %>%
  mutate(has_total_equal_number = any(number == total)) %>%
  filter(!has_total_equal_number) %>%
  pull(number) %>%
  unique()
```

**Answer:** `r format(invalid_number, scientific = FALSE)`

## Part 2

This part is very similar to part 1, except that we do not know in advance how many consecutive numbers we have to sum up to get to `r format(invalid_number, scientific = FALSE)`.

```{r}
df <- file %>%
  read_lines() %>%
  enframe(name = "number_id", value = "number") %>%
  mutate(number = parse_number(number))

# Try all combinations using between 2 and 50 numbers (should be ok)
max_length <- 2:50

df <- df %>%
  crossing(l = max_length) %>%
  arrange(number_id) %>%
  group_nest(l) %>%
  mutate(res = map2(data, l, function(data, l) {
    data %>%
      mutate(res = slide(number,
        ~ .x[1:l],
        .before = l,
        .complete = TRUE
      ))
  })) %>%
  unnest(res)

df <- df %>%
  filter(!map_lgl(res, is.null))

df %>% 
  head(100) %>% 
  rmarkdown::paged_table()
```

```{r}
df <- df %>%
  mutate(total = map2_dbl(res, l, function(data, l) {
    total <- combn(data, m = l) %>%
      t() %>%
      rowSums()

    return(total)
  }))
```

Find the case where the sum is equal to the answer found in part 1.

```{r}
df <- df %>%
  filter(total == invalid_number) %>%
  unnest(res)
```

```{r}
x <- df %>%
  pull(res)

x

range(x)
```

Sum up the minimum and maximum numbers.

**Answer:** `r sum(range(x))`
