---
title: "Day 03"
description: |
  Advent of code 2020 day 3
author:
  - name: Philippe Massicotte
date: 12-11-2020
output:
  distill::distill_article:
    self_contained: false
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(here)
```

## Context

https://adventofcode.com/2020/day/3

<small>
With the toboggan login problems resolved, you set off toward the airport. While travel by toboggan might be easy, it's certainly not safe: there's very minimal steering and the area is covered in trees. You'll need to see which angles will take you near the fewest trees.

Due to the local geology, trees in this area only grow on exact integer coordinates in a grid. You make a map (your puzzle input) of the open squares (.) and trees (#) you can see. For example:

```
..##.......
#...#...#..
.#....#..#.
..#.#...#.#
.#...##..#.
..#.##.....
.#.#.#....#
.#........#
#.##...#...
#...##....#
.#..#...#.#
```
</small>

## Read the input map

```{r read_input}
file <- here("data", "input", "day3.txt")

x <- read_lines(file)

m <- str_split(x, pattern = "") %>%
  do.call(rbind, .)

dim(m)

head(m)
```

# Part 1

## Create position vector

The rules for the first part are:

> The toboggan can only follow a few specific slopes (you opted for a cheaper model that prefers rational numbers); start by counting all the trees you would encounter for the slope right 3, down 1.

```{r}
start_i <- 1
start_j <- 1

delta_i <- 1
delta_j <- 3
```

## Duplicate the matrix

We need to have at least as many columns as rows.

```{r}
times <- ceiling(dim(m)[1] / dim(m)[2]) * delta_j
m <- matrix(rep(m, times), ncol = times * ncol(m), byrow = F)

dim(m)
```

```{r}
i <- seq((start_i + delta_i), nrow(m), by = delta_i) # lines
j <- seq((start_j + delta_j), ncol(m), by = delta_j) # columns

j <- j[1:length(i)]
```

## Count the number of trees

```{r}
object <- m[cbind(i, j)]
sum(str_count(object, "#"), na.rm = TRUE)
```

# Part 2

The second part is exactly the same as for part 1 but with different rules (or scenarios). 

```
Right 1, down 1.
Right 3, down 1. (This is the slope you already checked.)
Right 5, down 1.
Right 7, down 1.
Right 1, down 2.
```

Lets turn the previous code into a function.


```{r}
count_trees <- function(delta_i, delta_j, m, start_i, start_j) {
  times <- ceiling(dim(m)[1] / dim(m)[2]) * delta_j
  m <- matrix(rep(m, times), ncol = times * ncol(m), byrow = F)

  i <- seq((start_i + delta_i), nrow(m), by = delta_i) # lines
  j <- seq((start_j + delta_j), ncol(m), by = delta_j) # columns

  j <- j[1:length(i)]

  object <- m[cbind(i, j)]

  ntrees <- sum(str_count(object, "#"), na.rm = TRUE)

  return(ntrees)
}

```

Now we can define the three new scenarios:

```{r}
# Define the scenarios
right <- c(1, 3, 5, 7, 1)
down <- c(1, 1, 1, 1, 2)
start_i <- 1
start_j <- 1

file <- here("data", "input", "day3.txt")
x <- read_lines(file)
m <- str_split(x, pattern = "") %>%
  do.call(rbind, .)

trees <- pmap_int(
  list(down, right),
  count_trees,
  m = m,
  start_i = start_i,
  start_j = start_j
)

trees
```

And finally we need to multiply all the numbers.

```{r}
prod(trees)
```

